Overview and Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1)

### 2014-03-06 WSSSPE1_report/WSSSPE1_summary_report.tex
improve text

For the purpose of studying software case studies, we classify the software
projects discussed in the workshop in two broad categories. The first is
utility software, software tools that enable and/or facilitate the development
of other tools and techniques to carry out scientific work. This includes the
software developed to efficiently utilize new research infrastructures. The
second is scientific software, software that was developed with an aim to solve
a class of scientific problems.

These categories have some overlap, showing in some cases where case studies do
not cleanly fit into a single category.  Our classification is based on two
criteria: first, the character of a software project that dominates and where
it `seems' to be the best fit, and, second, the the workshop discussion session
where the paper about the project was discussed.

Our discussion of the case studies is aimed at understanding the practicalities of
the many points discussed in the workshop, and we hope to extract the best
practices and associate them with the projects. We discuss how the cases are
applicable to and have impacted user communities over their lifetime, isolating
similarities and differences between projects. We try to define the takeaways for a new
project and the lessons learned. We try to understand how the course of current and
future development projects might be altered to implement lessons learned.

\subsection{Utility Software}
Utility software potentially has the broadest community impact of the three
categories. This is especially relevant in the modern era of social computing.
In the context of this paper, a utility software project could be defined as
general purpose software applicable to one or many generic tasks and/or
enabling other software to run by providing a suitable environment. Examples
are collaborative development frameworks such as \emph{GitHub} and
\emph{Bitbucket}, workflow and scripting frameworks such as \emph{Galaxy},
\emph{Swift}, \emph{Globus} and \emph{VisTrails}, and visualization frameworks
such as \emph{VisIT}. Often projects start with a specific, small set of
requirements and users in mind but expand to provide more generic functionality
and appeal to a wider user community. Owing to its generic nature, utility
software finds a wider user community which in turn leads to a higher
participation in development. Consequently, the process becomes user-driven and
self-sustaining.

@@

%\subsection{Infrastructure software}
\todo{expand}
Utility software is often developed aimed at better utilizing particular, new
infrastructures and architectures, e.g., \emph{MVAPICH}, \emph{VisIT}.
Development in many cases is often a high risk/reward undertaking and explores
uncharted territories.

%
%\note{Dan: so is the only difference here that this category is aimed at new systems, where utility software is not?  If so, perhaps this should be incorporated into that subsection, so there would just be two subsections.  Something in the framing of this subsection doesn't feel right to me.  FL: I respectfully disagree the the only reason to develop frameworks is to better utilize (new or old) infrastructure. Another really important reason is to modularize development, making it much easier for an individual developer/user to integrate their own code. Ketan: I think in this section I wanted to bring up a class of software projects which are undertaken with the main aim of tackling new systems and architectures in mind. New infrastructures such as GPU, Cloud, fast IB, federal science networks, etc. were on my mind as I see many submissions around them in the workshop.}
%
Successful projects reap high rewards and have longer usage span. Long
requirement-development-test cycles.  Wide community support but tends to be
handled by specialists rather then end users. Promotes collaborations across
the breadth (e.g., different departments) and depth (e.g., stakeholders within
a department) of community, one of the key ingredient of a sustainable process.
Ferenbaugh~\cite{Ferenbaugh_WSSSPE} discusses the risks and challenges
associated with software projects dealing with new architectures. To leverage
the power of accelerators such as GPUs and MICs new code and libraries are
required. The experience of efforts as described in~\cite{Ferenbaugh_WSSSPE}
met with a limited success but nonetheless with many invaluable lessons were
learned about influential cultural and technical aspects in sustainable
software development practices.

@@

\subsection{Scientific software}
Scientific software consists primarily of special-purpose software that was
developed for a target use-case scenario or keeping some fixed/frozen
requirements in mind. Software projects pertaining to specific scientific
domain often tend to be in a niche and the user community tends to be small to
medium. Numerical accuracy and algorithmic optimization are some of the
paramount requirements of most scientific software. Advantages of a smaller
codebase and requirements include stability, ease of installation,
configurability and integration into larger systems. While the software can
stay stable and require relatively low maintenance, the responsibility is often
on the shoulders of a very few developers. Community participation in the
development tends to be low. Development is linear and simplistic with a
limited scope to follow software best practices. Domain science oriented
software is developed to solve  specific problems, e.g., \emph{DUNE},
\emph{PETSc}, \emph{MINRES-QLP}, \emph{FASTMath}. Sustainability of such
software is often a significant challenge. Many submissions reported the
software is considered a `byproduct' of the actual research.  Others contended
that the software was not the main funded part of their research. One exception
was \emph{Kitware}, which while being a software product specializing in the
scientific process, has a core focus of developing communities around software
processes. One example of this process is the development of the \emph{CMake}
build utility, which started out as a building tool for \emph{ITK} but grew to
become a generic build utility for C++ projects.

>>>>

For the purpose of studying software case studies, we classify the software
projects discussed in the workshop in two broad categories. First, the utility
software, comprising software tools that enable and/or facilitate the
development of other tools and techniques to carry out scientific work. This
includes the software developed to efficiently utilize new research
infrastructures.  Second, the scientific software, comprising software that was
developed with an aim to solve a class of scientific problems.

These categories have overlap in some cases where case studies do not cleanly
fit into a single category.  Our classification is based on two criteria:
first, the character of a software project that dominates and where it `seems'
to be the best fit, and, second, the workshop discussion session where the
paper about the project was discussed.

Our discussion is aimed at understanding the practicalities of the points
discussed in the workshop, and we hope to extract the best practices and
associate them with the projects. We discuss how the cases are applicable to
and have impacted user communities over their lifetime, isolating similarities
and differences between projects. We discuss the takeaways for a new
project and the lessons learned. We try to understand how the course of current
and future development projects might be altered to implement lessons learned.

\subsection{Utility Software}
Utility software has the potential of the broadest community impact of the two
categories. This is especially relevant in the modern era of social computing.
In the context of research, a utility software project could be defined as
general purpose software applicable to one or many generic tasks and/or
enabling other software to run by providing a suitable environment. Examples
are collaborative development frameworks such as \emph{GitHub} and
\emph{Bitbucket}, workflow and scripting frameworks such as \emph{Galaxy},
\emph{Swift}, \emph{Globus} and \emph{VisTrails}, and visualization frameworks
such as \emph{VisIT}. Often projects start with a specific, small set of
requirements and users in mind but expand to provide more generic functionality
and appeal to a wider user community. Owing to its generic nature, utility
software finds a wider user community which in turn leads to a higher
participation in development. Consequently, the process becomes user-driven and
self-sustaining.

@@

Utility software is often developed aimed at better utilizing particular, new
infrastructures and architectures, e.g., \emph{MVAPICH}, \emph{VisIT}.
Development in many cases is often a high risk/reward undertaking and explores
uncharted territories. 
%
%\note{Dan: so is the only difference here that this category is aimed at new systems, where utility software is not?  If so, perhaps this should be incorporated into that subsection, so there would just be two subsections.  Something in the framing of this subsection doesn't feel right to me.  FL: I respectfully disagree the the only reason to develop frameworks is to better utilize (new or old) infrastructure. Another really important reason is to modularize development, making it much easier for an individual developer/user to integrate their own code. Ketan: I think in this section I wanted to bring up a class of software projects which are undertaken with the main aim of tackling new systems and architectures in mind. New infrastructures such as GPU, Cloud, fast IB, federal science networks, etc. were on my mind as I see many submissions around them in the workshop.}
%
Successful projects reap high rewards and have longer usage span. Long
requirement-development-test cycles.  Wide community support but tends to be
handled by specialists rather then end users. Promotes collaborations across
the breadth (e.g., different departments) and depth (e.g., stakeholders within
a department) of community, one of the key ingredient of a sustainable process.
Ferenbaugh~\cite{Ferenbaugh_WSSSPE} discusses the risks and challenges
associated with software projects dealing with new architectures. To leverage
the power of accelerators such as GPUs and MICs new code and libraries are
required. The experience of efforts as described in~\cite{Ferenbaugh_WSSSPE}
met with a limited success but nonetheless with many invaluable lessons were
learned about influential cultural and technical aspects in sustainable
software development practices.

@@

\subsection{Scientific software}
Scientific software consists primarily of special-purpose software that was
developed for a target use-case scenario or fixed/frozen requirements in mind.
Software projects pertaining to specific scientific domain often tend to be in
a niche and the user community tends to be small to medium. Specific
requirements such as numerical accuracy and algorithmic optimization are some
of the paramount requirements of most scientific software. The resulting
codebase is relatiely small. A smaller codebase and fixed requirements result
in stability, ease of installation, and configurability. Many such projects are
treated as libraries used into larger systems such as some of the utility
software discussed in the previous section. While the software can stay stable
and require relatively low maintenance, the responsibility is often on the
shoulders of a very few developers. Community participation in the development
tends to be low. Development is linear and simplistic with a limited scope to
follow software best practices. Domain science oriented software is developed
to solve  specific problems, e.g., \emph{DUNE}, \emph{PETSc},
\emph{MINRES-QLP}, \emph{FASTMath}. Sustainability of such software is often a
significant challenge. Many submissions reported the software is considered a
`byproduct' of the actual research.  Others contended that the software was not
the main funded part of their research. One exception was \emph{Kitware}, which
while being a software product specializing in the scientific process, has a
core focus of developing communities around software processes. One example of
this process is the development of the \emph{CMake} build utility, which
started out as a building tool for \emph{ITK} but grew to become a generic
build utility for C++ projects.
